Looking for a Signal in the Noise: Revisiting Obesity and the Bacterial Microbiome

### 2016-02-23 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
adding Pat's suggestions to manuscript

*******
##### **Abstract:**
**Rationale:** The hypothesis that changes in the microbiome and obesity are associated has been widely accepted.  However, two recent reports that re-examined existing data suggest that acceptance of this hypothesis is unwarranted.  These studies, however, studied the original studies in isolation from each other. We hypothesized that by statistically pooling the individual results, the statistical power to identify a significant association would increase.  In this meta analysis, we tested the significance of the association between the microbiome and obesity by increasing the number of data sets examined and investigating the effect of pooling the results across the individual studies.
**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did X, Y, Z (fill in criteria), a total of 9 data sets were identified. We applied three analysis piplines to the data from each study. First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and XXXX with the subjects' body mass index (BMI) and obesity category.  Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each studies.  Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.  [aren't these two points the same thing? didn't you have to normalize the data to pool the results? Also, what about the RF stuff?]
**Results:** There were no significant differences (P > 0.05) between the relative abundance of Bacteroidetes, Firmicutes, or thier ratio in non-obese and obese individuals. The Shannon diversity, OTU richness, relative risk, or oveall community dissimilarity was significant in at least one study (P<0.05).  When we pooled the results across studies, there was a signifcant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)) and B/F ratio (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)). [RF results?]
**Conclusion:** Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.  [this repeats the previous sentence. why not say something about how the approach of pooling results offers an opportunity to test the generalizability of microbiome results by effectively increasing sample size or something?]

*******
##### **Introduction:**  
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].  This number increases to approximately 35% in adults (aged 20 or older) and these statisitics have seen little change since 2003 [@ogden_prevalence_2014].  Traditionally the body mass index (BMI) has been used as the traditional method of classifying indviduals as non-obese or obese [@lichtash_body_2013].  However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures of traditionally associated with obesity [@lichtash_body_2013].  Recently, there has been a lot of interest in the bacteiral microbiome and it's potential ability to modulate obesity [@brahe_can_2016; @dror_microbiota_2016].  If this modulation potential is true then the bacterial microbiome could have a enormous impact and role to play in the future treatment of obesity and helping to stem the rising obesity epidemic.

@@

The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing applying a more sytematic approach than has been used by previous studies in this area.  Additionaly, this study will include both seperate and pooled analysis to assess the bacterial microbiome and obesity. By including these two components this study will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.  In particular, the goal of this study is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.

Although there have been numerous RT-PCR and qPCR studies directed at examining the bacterial microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].  The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.  The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454, Illumina MiSeq, and Illumina HiSeq.  For this report we limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.  We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.

*******
##### **Methods:**  
*Literature Review and Study Inclusion*
<br><br>
We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010]. A detailed description of the literature seach and the exact formula used for the initial PubMed search can be found in the online supplement along with the PRISMA flow diagram [Figure 1].  In brief, a total of 196 studies made it through the initial identification of studies.  Screening removed a total of 184 studies leaving a total of 12 eligible studies for this meta-analysis.  Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement.  A total of 10 studies were ultimately deemed eligible for qualitative synthesis.  One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs. A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012].  One of these studies (Baxter) has not been published at the time of this writing.

*Subject Demographics and Study Outcomes*
<br><br>
The overall demographics for indivdiuals used in this meta-analysis split by their respective study can be found in table 1.  A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2.  There is a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.  However, a total of three manuscripts found no correlations between the bacterial microbiome and obesity.  This is in addition to the two re-analysis studies published in 2014.  The cutoffs used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), obese (BMI greater than or equal to 30).  For analysis involving obese and non-obese groups the cutoffs used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).

*Sequence Analysis Pipeline*
<br><br>
All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.  A total of 6/9 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.  A total of 2/9 [@goodrich_human_2014] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.  The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.  A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1-S3]. For the studies that utilized 16S rRNA gene sequencing three studies used primers targeting the V1-V3 region [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], two studies used the V4 region [@goodrich_human_2014], and one study used the V1-V2 region [@wu_linking_2011].  Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009].  A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.

*Data Analysis*
<br><br>
All analysis was performed using R version 3.2.2 and R studio version 0.99.491.  The overall analysis was split into three general categories.  The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.  The main associations with BMI that were explored in each seperate data set were Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).  A two-way T-test was performed for comparison for non-obese versus obese.  An ANOVA was used to investigate all BMI categories (if available) with a Tukey post hoc test. Third, a Pearson test for linear correlation between the variables was used.  Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.  The R package AUCRF (v1.1) was used for this.
<br><br>
The second category involved pooling all the data from all the studies together to asses whether the combined direction of all the studies increased an indivdiual's RR for being obese.  The two measurements that were investigated in detail for this part were Shannon diversity and the B/F ratio.  To generate the RR the median value of each study was taken and the number of those that were non-obese or obese in each group was recorded and a Fisher exact-test was performed.  The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.  To obtain the pooled RR and confidence intervals (CI) the metafor (v1.9-8) package available for the R statistics software was utilized.  Finally, to assess potential bias we also analyzed the two pooled analysis using a funnel plot. 
<br><br>
The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between BMI and Shannon diversity, Bacteroidetes, Firmicutes, and B/F ratio.  Using the Zscore values a two-way t-test was used to investigate if there was a difference between non-obese and obese individuals.  Differences in BMI categories (if data was available) was investigated using ANOVA with a tukey post hoc test.  Linear correlations between these variables and BMI were explored using the Pearson test.  Where appropriate the Bonferroni correction was applied for multiple comparisons.  In all other cases a P-value under 0.05 was considered to be significant.

@@

We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].  Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 3 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.  There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.  The full summary of the exact P-values for each respective study and variable test can be found in table 3.

When investigating how well the OTUs in each study were able to classify obese and non-obese individuals the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.  The total range of the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).  The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.  Perhaps surprisingly there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.  The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/9 studies), an OTU that could be classified to the family Lachnospiraceae (5/9 studies), and the Shannon diversity metric (4/9 studies).

@@

When the Shannon diversity RR, all studies but Wu et al had those with lower Shannon diversity then the median as being at higher risk for obesity [Figure 2].  This was confirmed with the pooled analysis showing an increased RR of 1.19 (CI 1.04, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0139).  For the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese indivdiuals than those with a high B/F ratio [Figure 3].  A notable exception to this was the HMP data set and the Baxter data set [Figure 3].  The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0296).

Using a funnel plot to asses the bias of the 9 data sets used in this meta-analysis we can observe that there does not seem to be any significant  bias [Figure 4].  Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

We also investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC performed from each study performed when analyzed with RR [Figure S4].  There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.98 (CI 0.63, 1.53), P-value = 0.9225).  Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].

@@

Using this approach we found that there was a significant difference in the normalized Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].  Further there was also a significant difference in the normalized Shannon diversity variance when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].  Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).  With respect to the B/F ratio using the ZScore normalization we found that there was no significant difference between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].  Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].

@@

This study helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.  Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a bacterial microbiome difference between non-obese and obese individuals.  However, neither study really made an attempt to pool the existing data together to try and harness the additional power that this would give.  Here we perform an extensive literature review of the existing studies on the bacterial microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.  We analyze the data using three different approaches.  The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses ZScore normalizations and does not use RR.

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the bacterial microbiome and obestiy could be observed.  However, the other two approaches using the pooled data was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].  For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.  Taken together this analysis provides a preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.  However, our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since both methods of pooled analysis found Shannon diversity significant while only one did for the B/F ratio.

This study does have a few limitations that are worth mentioning.  First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].  Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].  Although the asain population in the studies analyzed is very small [Table 1] it is possible that this may contribute to some of the noise in the data.  However, for the measures, specifically the RR, there does not seem to be a study bias associated with this [Figure 4].  We are also limited in our BMI classifications by what the previous studies published and performed.  In general, not enough studies have been done investigating the bacterial microbiome in the context of other markers for obesity or for different ethnicities to be able to perform the analysis that we did.  Thus as more specific studies are published that address ethnicity and better markers of obesity it is possible that the noise associated with these measurements will decrease.

Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the bacterial microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.  However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.

Third, this meta-analysis and review of the data to date is still rather preliminary.  Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study.  It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the bacterial microbiome within humans.

@@

Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in the effort to find a correlation between the bacterial microbiome and obesity.  In doing so it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.  However, it should be cautioned that there are still many aspects of this analysis that could be improved upon and it truly is only a first attempt at trying to summerize and analyze the pooled data that is available.  We hope that this study helps to add important information to the on-going dialogue of the bacterial microbiome's role in obesity.

@@

>>>>

*******
##### **Abstract:**
**Rationale:** The hypothesis that changes in the microbiome and obesity are associated has been widely accepted.  However, two recent reports that re-examined existing data suggest that acceptance of this hypothesis is unwarranted.  These studies, however, studied the original studies in isolation from each other. We hypothesized that by statistically pooling the individual results, the statistical power to identify a significant association would increase.  In this meta analysis, we tested the significance of the association between the microbiome and obesity by increasing the number of data sets examined and investigating the effect of pooling the results across the individual studies.

**Methods:** We identified 196 studies that tested the microbiome-obesity association hypothesis; however when we limited the studies to those that did microbiome sequencing, had publicly available sequences, and were either willing to share the BMI information for every individual or were already published, a total of 9 data sets were identified. We applied three analysis piplines to the data from each study. First, we tested the association between the community diversity, Bacteroides to Firmicutes ratio (B/F), and relative risk (RR) with the subjects' body mass index (BMI) and obesity category.  We also used random forest to identify variables for classification of obese and non-obese indivdiuals.  Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each studies.  Finally, data normalization was used to investigate all data sets together for relationships between the microbiome and obesity.

**Results:** There were no significant differences (P > 0.05) between the relative abundance of Bacteroidetes, Firmicutes, or thier ratio in non-obese and obese individuals. The Shannon diversity, OTU richness, relative risk, or oveall community dissimilarity was significant in at least one study (P<0.05).  Overall, the random forest results for Area Under the Curve (AUC) was 0.7553 +/- 0.0888 and 7/9 studies contained an OTU that could be classified to the Ruminococcaceae family.  When we pooled the results across studies, there was a signifcant difference in RR of obesity based on Shannon diversity (RR = 1.19, CI = 1.04, 1.36, (P-value = 0.0139)) and B/F ratio (RR = 0.86, CI = 0.75, 0.99, (P-value = 0.0296)).

**Conclusion:** We show that by pooling the results of many indivdiual studies we can test the generalizability of the bacterial microbiome.  Our meta-analysis supports the  hypothesis that there is a difference between non-obese and obese individuals in Shannon diversity and B/F ratio, based on the currently avialable bacterial microbiome data included in this meta-analysis.

*******
##### **Introduction:**  
Obesity is a growing worldwide health concern with approximately 20% of the youth (aged 2-19) in the United States classified as either overweight or obese [@ogden_prevalence_2014].  This number increases to approximately 35% in adults (aged 20 or older) and these statisitics have seen little change since 2003 [@ogden_prevalence_2014].  Traditionally the body mass index (BMI) has been used as the traditional method of classifying individuals as non-obese or obese [@lichtash_body_2013].  However, other methods do exist, such as body adiposity index (BAI) but do not always correlate with important molecular measures traditionally associated with obesity [@lichtash_body_2013].  Recently, there has been a lot of interest in the bacteiral microbiome and it's potential ability to modulate obesity [@brahe_can_2016; @dror_microbiota_2016].  If this modulation potential is true then the bacterial microbiome could have an enormous impact and role to play in the future treatment of obesity and in helping to stem the rising obesity epidemic.

@@

Although there have been numerous RT-PCR and qPCR studies directed at examining the bacterial microbiome and obesity [@guo_development_2008; @schwiertz_microbiota_2010].  The main method of choice to obtain the most complete picture of the overall microbial community is through 16S sequencing of a specific hypervariable region on the 16S rRNA gene.  The sequencers used to complete this task have varied depending on research group with the most popular being the Roche 454 and Illumina MiSeq.  A priori we decided that for this report we would limit ourselves to analyzing studies were 16S sequencing data is avaliable along with BMI information.  We also analyzed every study with a uniform approach utilizing the mothur software package for sequence processing and clean up and R statistical software for analysis.

The purpose of this study is to perform an extensive meta-analysis of the bacterial microbiome and obesity by analyzing and applying a more sytematic approach than has been used by previous studies in this area.  Additionaly, we will include both seperate and pooled analysis to assess the bacterial microbiome and obesity. By including these two components we will build on the previous two re-analysis studies and provide a stronger assesment of the direction of effect or lack there of on the bacterial microbiome and obesity.  In particular, our goal is to provide additional clarity on the role of both bacterial diversity and the Bacteroidetes/Firmicutes ratio impact on obesity.

*******
##### **Methods:**  
*Literature Review and Study Inclusion*

We attempted to stick to the original PRISMA guidelines for performing a meta-analysis as published by Moher et al [-@moher_preferred_2010]. A detailed description of the literature seach and the exact formula we used for the initial PubMed search can be found in the online supplement.  In brief, a total of 196 studies made it through the initial identification of studies.  Screening removed a total of 184 studies leaving a total of 12 eligible studies for this meta-analysis.  Upon a full reading of these specific manuscripts 2 were exculded and the reasons can be found in the online supplement.  A total of 10 studies were ultimately deemed eligible for qualitative synthesis [Figure 1].  One final study was removed from the quantitative analysis since no individual in the study was found to be able to be classified in the obese category based on the classical cutoffs. A total of 3 studies were from records identified by database searching [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014] and the other 6 were from records identified by other sources [@ross_16s_2015; @arumugam_enterotypes_2011; @goodrich_human_2014; @wu_linking_2011; @conlan_species-level_2012].  One of these studies (Baxter) has not been published at the time of this writing.

*Subject Demographics and Study Outcomes*

The overall demographics for individuals used in this meta-analysis split by their respective study can be found in table 1.  A summary of the study type, study population, whether the data set is published, and the outcomes reported can be found in table 2.  We found a large variation in reported outcomes with two studies finding bacterial community associations with obesity and three studies finding some association between either Firmicutes, Bacteroidetes, or the Bacteroidetes/Firmicutes ratio and obesity.  However, a total of three manuscripts found no correlations between the bacterial microbiome and obesity.  This is in addition to the two re-analysis studies published in 2014.  The cutoffs we used in this study for the BMI groups were lean (BMI less than or equal to 24), overweight (BMI was greater than 24 but less than 30), and obese (BMI greater than or equal to 30).  For analysis involving obese and non-obese groups the cutoffs we used were a BMI less than 30 (lean/non-obese) and a BMI of equal to or greater than 30 (obese).

*Sequence Analysis Pipeline*

All data was publically available and the sequence data was downloaded from the NCBI Sequence Read Archive, MG-RAST, or the European Nucleotide Archive.  A total of 6/9 [@turnbaugh_core_2009; @ross_16s_2015; @zupancic_analysis_2012; @wu_linking_2011; @escobar_gut_2014; @conlan_species-level_2012] studies were processed by staying as close as possible to the mothur 454 standard operating procedure, which can be found at http://www.mothur.org/wiki/454_SOP.  A total of 2/9 [@goodrich_human_2014] of the studies were processed by staying as close as possible to the mothur MiSeq standard operating procedure found at http://www.mothur.org/wiki/MiSeq_SOP.  The final data set [@arumugam_enterotypes_2011] that was used took advantage of the Metaphlan2 package [@truong_metaphlan2_2015] to generate the necessary data for this analysis.  A comparison of the similarity of Metaphlan2 to 16S sequencing using the Human Microbiome Project (HMP) data set can be found in the online supplement [Figure S1-S3]. For the studies that utilized 16S rRNA gene sequencing three studies used primers targeting the V1-V3 region [@ross_16s_2015; @escobar_gut_2014; @zupancic_analysis_2012], two studies used the V4 region [@goodrich_human_2014], and one study used the V1-V2 region [@wu_linking_2011].  Where there were two choices of data sets to use we used the V3-V5 data of the HMP [@conlan_species-level_2012] and the V2 data from the Turnbaugh study [-@turnbaugh_core_2009].  A detailed walkthrough of the entire analysis pipeline can be found at https://github.com/SchlossLab/Sze_ObesityMicrobiomeMeta_PeerJ_2016/.

*Data Analysis*

We performed the analysis using R version 3.2.2 and R studio version 0.99.491.  We split the overall analysis into three general categories.  The first category involved following the same method employed by Finucane et al. and Walters et al. [-@finucane_taxonomic_2014; -@walters_meta-analyses_2014] in which each study was re-analyzed seperately for associations with BMI.  The main associations that we explored for BMI in each seperate data set were the Bacteroidetes phyla, Firmicutes phyla, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, NMDS visualization of Bray-Curtis distance matrix PERMANOVA, and relative risk (RR).  A two-way T-test was performed for comparison for non-obese versus obese.  Finally, Random forest analysis to identify the best variables for classification of obese and non-obese was performed to look if there were similarities between data sets that we may have missed.  The R package AUCRF (v1.1) was used for this.

For the second category of analysis we pooled all the data from each seperate study together to asses whether the combined direction of all the studies increased an individual's RR for being obese.  We investigated both Shannon diversity and the B/F ratio in detail for this part.  In order to generate the RR we used the median value of each study and the number of those that were non-obese or obese in each group (above or below the median respectively) was recorded and a Fisher exact-test was performed.  The relative risk from the the Fisher exact-test was obtained using the epiR (v0.9-6.9) package in R.  We obtained the pooled RR and confidence intervals (CI) from the metafor (v1.9-8) package in R.  Finally, we also assessed the potential bias in the two pooled analysis by using a funnel plot.

The third and final category involved using  Zscore normalizations to pool all the data together to investigate the correlation between either BMI and Shannon diversity or BMI and the B/F ratio.  We used a two-way t-test on the Zscore values to investigate if there was a difference between non-obese and obese individuals.  In order to asses differences between the BMI categories (if data was available) we used ANOVA with a tukey post hoc test.  Where appropriate we applied a Bonferroni correction for multiple comparisons.  In all other cases a P-value under 0.05 was considered to be significant.

@@

We used the categorical variable of obese or non-obese to analyze each data set seperately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicute (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014].  Using a P-value cutoff for significance as less than 0.05 we found a total of 1 significant result for Shannon diversity, 3 significant results for OTU richness, and 1 signficant result for the Bray-Curtis distance matrix.  There were no significant results for Bacteroidetes, Firmicutes, or the B/F ratio.  The full summary of the exact P-values for each respective study and variable tested can be found in table 3.

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm.  We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was 0.7553 +/- 0.0888.  We found that the total range for the total number of variables used for the classification ranged from 3 (Ross and Escobar) to 32 (Zupancic).  The total number of trees and node size was standardized to 1000 and 20 respectively for each data set.  One aspect we found surprising was that there was very little overlap in the variables used by the random forest algorithm for the classification of obese and non-obese individuals by study.  The three most common variables was for an OTU that could be classified to the family Ruminococcaceae (7/9 studies), an OTU that could be classified to the family Lachnospiraceae (5/9 studies), and the Shannon diversity metric (4/9 studies).

@@

We next moved on to the pooled meta-analysis.  For the Shannon diversity RR, all studies but Wu et al had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 2].  This was confirmed with the pooled analysis showing an increased RR of 1.19 (CI 1.04, 1.36) for obesity in the lower then median Shannon diversity group (P-value = 0.0139).  When investigated the B/F ratio RR most studies showed that a low B/F ratio tended to have less obese individuals than those with a high B/F ratio [Figure 3].  A notable exception to this was the HMP data set and the Baxter data set [Figure 3].  The pooled RR was 0.86 (CI 0.75, 0.99) suggesting that a low B/F ratio does indeed have less obese individuals (P-value = 0.0296).

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis.  From the graph we can observe that there does not seem to be any significant  bias [Figure 4].  Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

The final forest plot that we used for this analysis investigated how the different OTUs, classified to Ruminococcaceae, that were important for the classification of obese and non-obese based on the OOB AUC in the the seperate study analysis looked when pooled together to assess the RR [Figure S4].  There was no significant increase in RR based on the Ruminococcaceae OTU (RR = 0.98 (CI 0.63, 1.53), P-value = 0.9225).  Thus it would seem that although Ruminococcaceae is important for classification the directionality is study dependent [Figure S4].

@@

Using the third approach with Zscore normalization we found that there was a significant difference in the Shannon diversity between the obese and non-obese group (P-value = 0.00493) [Figure 5A].  Further there was also a significant difference in this normalized Shannon diversity value when the data was split between the three BMI groups (P-value = 0.03826) [Figure 5B].  Using a Tukey post hoc test this difference was driven by the overweight versus the obese group (P-value = 0.03945).  However, in contrast to this result we found that the B/F ratio, using the ZScore normalization, was not significant different between obese and non-obese indivdiuals (P-value = 0.4516) [Figure 5C].  Additionally, we did not observe a difference when the indivdiuals were seperated by BMI groups (P-value = 0.3317) [Figure 5D].

@@

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific bacterial microbiome variables that are influenced by obesity.  Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a bacterial microbiome difference between non-obese and obese individuals.  However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give.  Here we perform an extensive literature review of the existing studies on the bacterial microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria.  We analyze the data using three different approaches.  The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach also takes a meta-analysis approach but uses ZScore normalizations and does not use RR.

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the bacterial microbiome and obestiy could be observed.  However, we found that the other two approaches, that used the pooled data, was able to identify a correlation between measures of the bacterial microbiome and obesity [Figure 2 & 3].  For Shannon diversity both approaches using the pooled data found a correlation while only one of the approaches for the B/F ratio found a correlation.  Taken together we believe that this analysis provides a strong preliminary finding that there is a small but detectable difference between non-obese and obese individuals with respect to both Shannon diversity and the B/F ratio.  However, it should be noted that our analysis would indicate that the signal for Shannon diversity may be more robust than that of the B/F ratio since the two different pooled analysis approaches found Shannon diversity significant while only one did for the B/F ratio.

We do acknowledge that this study does have a few limitations that are worth mentioning.  First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010].  Of particular note, the asian population has a seperate set of guidelines for obesity [@who_expert_consultation_appropriate_2004].  Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data.  However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4].  Additionally, we are also limited in our BMI classifications by what the previous studies published and performed.  In general, not enough studies have been done to investigate the bacterial microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did.  Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.

Second, although we make attempts to normalize the data for the pooled analysis so that we can compare the bacterial microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.  However, what our analysis shows is that despite these differences, we are still able to identify an overall signal from the data.    
Third, this meta-analysis and review of the data to date is still rather preliminary.  Only a total of 1542 individuals were included in the pooled analysis which is relatively small compared to other meta-analysis in non-microbiome fields of study.  It will be necessary to continuously update this analysis, as more studies become available, in order to obtain a more accurate estimate of the true effect that obesity has on the bacterial microbiome within humans.

@@

Despite the many problems that could add more noise to the measurements used, this study provides a reasonable first pass attempt to pool this data together in an effort to find a correlation between the bacterial microbiome and obesity.  In doing so we believe that it provides a tanatalizing result that despite the large differences between studies it is still possible to detect a correlation between the bacterial microbiome and obesity.  However, we emphaszie that it should still be viewed with caution since there are still many aspects of this analysis that could be improved upon.  It truly is only a first attempt at trying to summerize and analyze the pooled data that is available.  We hope that this study helps to add important information and clarity to the on-going dialogue of the bacterial microbiome's role in obesity.

@@

```{r Data_Baxter, echo=FALSE, message=FALSE, warning=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

# Reading in the necessary Data
setwd("C:/Users/marc/Desktop/obesity2/data1")
demographics <- read.csv("demographics.v2.csv")
microbiome <- read.csv("data1.subsample.otus.csv")
phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
taxonomy <- read.csv("data1.taxonomy.csv")

# Minor modifications and Rownames adustments of data tables
rownames(demographics) <- demographics[,1]
rownames(microbiome) <- microbiome[,2]
demographics <- demographics[,-1]
microbiome <- microbiome[,-2]
test.samples <- rownames(demographics)
microbiome <- microbiome[,-c(1:2)]

# Create Table for phyla information
phylogenetic.info <- phylogenetic.info[-c(1:12),]
phyla <- which(phylogenetic.info$taxlevel == 2)
phyla.table <- phylogenetic.info[phyla, ]
phyla.table <- phyla.table[,-c(1:2,4:5)]
phyla.names <- as.character(phyla.table[,1])
phyla.table <- phyla.table[,-1]
phyla.table <- as.data.frame(t(phyla.table))
colnames(phyla.table) <- phyla.names
rownames(phyla.table) <- rownames(demographics)

#Add phyla together that are not very abundant and delete them from the table
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]
